---
title: "Custom ODE solvers in Stan"
author: "Juho Timonen"
date: "3/11/2020"
output: 
  html_vignette:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, results=FALSE, warning=FALSE, message=FALSE}
require(rstan)
require(ggplot2)
require(bayesplot)
source('plotting.R')
rstan_options(auto_write = TRUE)
color_scheme_set("red")
```

## 1. Defining a toy system
Here we study a Lotka-Volterra system defined as
$$
\begin{cases}
\dot{x_1} &= x_1 - \theta_1 x_1 x_2 \\
\dot{x_2} &= \theta_1 x_1 x_2- \theta_2 x_2
\end{cases}
$$
with $x_1(0) = 1$ and $x_2(0) = 2$. We create a Stan model.
```{r model1, cache=TRUE}
# The ODE system and integrator are specified in these files
odeint_file <- 'stan/odefun_lotka-volterra.txt'
odefun_file <- 'stan/odeint_rk45-boost.txt'

# This creates a complete stan program code from three parts
# by adding a functions block to a base file
create_stan_code <- function(odeint_file, odefun_file, base_file = 'stan/base.txt'){
  base <- readLines(base_file)
  base <- base[(12+1):length(base)] # remove comment rows from base
  fun1 <- readLines(odeint_file)
  fun2 <- readLines(odefun_file)
  code <- paste(c('functions{', fun1, fun2, '}', base), collapse="\n")
  return(code)
}

# Create model code and compile model
code  <- create_stan_code(odeint_file, odefun_file)
model <- stan_model(model_code = code)
```

We can expose its functions to R.
```{r expose}
# Expose odefun() and odeint() to R
expose_stan_functions(model)
```

The model code looks like this:
```{r code1, echo = FALSE}
cat(code)
```

## 2. Simulating data
We generate data on which we can test our inference.
```{r sim, fig.width=7.2, fig.height=3.5}
# Simulation setup
set.seed(123)
t_data <- seq(0.5,12,by=0.5)
y0     <- c(1,2)     # initial state
theta  <- c(1,1)     # true parameter values
sigma  <- 0.2        # noise std
N      <- length(t_data)
D      <- length(y0)
P      <- length(theta)

# Convenience wrapper for a function obtained by expose_stan_functions
odeint_wrap <- function(odeint_method, y0, t0, t, theta){
  x_r <- x_i <- rep(0, 0)
  Y   <- odeint_method(y0, t0, t, theta, x_r, x_i)
  Y   <- matrix(unlist(Y), length(t), length(y0), byrow = TRUE)
  return(Y)
}

# Create and plot data
Y_data <- odeint_wrap(odeint, y0, 0, t_data, theta) + matrix(rnorm(n=N*D, sd=sigma), N, D)
plot_data(t_data, Y_data)
```

## 3. Inference using the built-in RK45 method

Here we fit a model where ODEs are solved using the `integrate_ode_rk45` which is built into Stan and uses the Boost implementation of the RK45 (or Dormand–Prince) method.

### 3.1 Fitting the model
```{r fit1, fig.width=12, fig.height=6, cache=TRUE}
# Formats data for Stan
create_stan_data <- function(y0, t_data, Y_data, P){
  N <- dim(Y_data)[1]
  D <- dim(Y_data)[2]
  out <- list(N=N, D=D, P=P, y0=y0, t0=0, t=t_data, y=Y_data)
  return(out)
}

# Fit model
stan_seed <- 321
data <- create_stan_data(y0, t_data, Y_data, P)
fit  <- sampling(model, data, seed = stan_seed, refresh=0, iter=1000, chains=3)
print(fit)
```

We visualize the 4 chains
```{r diag1, fig.width=7.2, fig.height=5}
# Diagnose
plot_chains(fit, ncol=2, n_warmup=500) + theme(text = element_text(size=14))
```

We see that one chain was much slower than others and spent much of the warmup time doing ode solves with parameters that give a very low log posterior probability (`lp__`).
```{r time1}
print(get_elapsed_time(fit))
```

### 3.2 Visualizing solutions
We visualize the ODE solutions against data.
```{r vis1, fig.width=7.2, fig.height=3.5}
# Takes subsamples of theta
subsample_theta <- function(fit, n_samples){
  ext <- rstan::extract(fit, pars=c('theta'), inc_warmup=TRUE, permuted=FALSE)
  samples <- rbind(ext[,1,], ext[,2,], ext[,3,])
  samples <- samples[sample(nrow(samples)),]
  samples <- samples[1:n_samples,]
  return(samples)
}

# Obtains solutions on a smoother time grid
solutions <- function(data, odeint_method, THETA, t){
  K  <- dim(THETA)[1]
  Y_hat <- array(0, c(K, length(t), length(data$y0)))
  for(k in 1:K){
    theta <- THETA[k, ]
    y_hat <- odeint_wrap(odeint_method, data$y0, 0, t, theta)
    Y_hat[k,,] <- y_hat
  }
  return(Y_hat)
}

# Solve at t_hat and plot
h     <- 0.1
t_hat <- seq(h, 15, by=h)
THETA <- subsample_theta(fit, 100)
Y_hat <- solutions(data, odeint, THETA, t_hat)
plot_solutions(t_data, Y_data, t_hat, Y_hat, alpha=0.3)
```

We plotted 100 randomly taken solutions including warmup.

## 4. Custom solver
We now specify a different solver in `odefun_file`, create new model and expose the new `odeint` function to R.

```{r model2, cache=FALSE}
# The ODE system and integrator are specified in these files
odeint_file <- 'stan/odefun_lotka-volterra.txt'
odefun_file <- 'stan/odeint_my-rk4.txt'

# This creates a complete stan program code from three parts
# by adding a functions block to a base file
create_stan_code <- function(odeint_file, odefun_file, base_file = 'stan/base.txt'){
  base <- readLines(base_file)
  base <- base[(12+1):length(base)] # remove comment rows from base
  fun1 <- readLines(odeint_file)
  fun2 <- readLines(odefun_file)
  code <- paste(c('functions{', fun1, fun2, '}', base), collapse="\n")
  return(code)
}

# Create model code and compile model
code  <- create_stan_code(odeint_file, odefun_file)
model <- stan_model(model_code = code)
```

We can expose its functions to R.
```{r expose2}
# Expose odefun() and odeint() to R
expose_stan_functions(model)
```

The model code looks like this:
```{r code2, echo = FALSE}
cat(code)
```

