---
title: "Custom ODE solvers in Stan - Part 2"
author: "Juho Timonen"
date: "3/12/2020"
output: 
  html_vignette:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The R functions and `.stan` files used in this vignette can be found [here](https://github.com/jtimonen/odeint-stan).

```{r libs, results=FALSE, warning=FALSE, message=FALSE}
require(rstan)
source('functions.R')
rstan_options(auto_write = TRUE)
color_scheme_set("red")
```

```{r model, message=FALSE, cache=TRUE}
model <- stan_model('stan/main.stan')
expose_stan_functions(model)
```

```{r expose, results=FALSE, warning=FALSE, message=FALSE}
expose_stan_functions(model)
```

## 1. Simulating data
We generate data using the system defined in [Part 1]('https://jtimonen.github.io/odeint-stan/part1.html'). Noisy observations of **x** are denoted by **y**.
```{r sim, fig.width=7.2, fig.height=3.5}
# Simulation setup
set.seed(27342)
t0     <- 0
y0     <- c(1,2)
t_data <- seq(0.5,12,by=0.5)
theta  <- c(1,1)     # true parameter values
sigma  <- 0.2        # noise std
N      <- length(t_data)
D      <- length(y0)
P      <- length(theta)

# Create and plot data
noise <- matrix(rnorm(n=N*D, sd=sigma), N, D)
Y_data <- odeint(rk45_boost, y0, t0, t_data, theta) + noise
plot_data(t_data, Y_data)
```

## 2. Inference using the built-in RK45 method

Here we fit a model where ODEs are solved using the `integrate_ode_rk45` which is built into Stan and uses the Boost implementation of the RK45 (or Dormand–Prince) method.

### 2.1 Fitting the model
```{r fit1, fig.width=12, fig.height=6, cache=TRUE}
stan_seed <- 243
iter <- 2000
chains <- 6
data <- create_stan_data(y0, t_data, Y_data, P, "rk45_boost",
                         abs_tol=1e-10, rel_tol=1e-10)
fit  <- sampling(model, data, seed = stan_seed, refresh=0, iter=iter, chains=chains)
print(fit)
```

We visualize the chains
```{r diag1, fig.width=7.2, fig.height=5}
plot_chains(fit, ncol=2)
```

```{r time1}
print(get_elapsed_time(fit))
```

### 2.2 Visualizing solutions
We visualize the ODE solutions against data.
```{r vis1, fig.width=7.2, fig.height=3.5}
t_hat <- seq(0.05, 15, by=0.05)
THETA <- subsample_theta(fit, 100)
Y_hat <- solutions(data, rk45_boost, THETA, t_hat)
plot_solutions(t_data, Y_data, t_hat, Y_hat, alpha=0.3)
```

We plotted 100 randomly taken solutions, some of which are from warmup.

## 3. Custom RK4 method

Now we fit a model where ODEs are solved using our `rk4` solver with 4 steps per output time point.

### 3.1 Fitting the model
```{r fit2, fig.width=12, fig.height=6, cache=TRUE}
data <- create_stan_data(y0, t_data, Y_data, P, "rk4",
                         n_steps_per_timepoint = 4)
fit  <- sampling(model, data, seed = stan_seed, refresh=0, iter=iter, 
                 chains=chains)
print(fit)
```

We visualize the chains
```{r diag2, fig.width=7.2, fig.height=5}
plot_chains(fit, ncol=2)
```

```{r time2}
print(get_elapsed_time(fit))
```

### 3.2 Visualizing solutions
We visualize the ODE solutions against data.
```{r vis2, fig.width=7.2, fig.height=3.5}
THETA <- subsample_theta(fit, 100)
Y_hat <- solutions(data, rk4, THETA, t_hat, n_steps_per_timepoint=4)
plot_solutions(t_data, Y_data, t_hat, Y_hat, alpha=0.3)
```

We again plotted 100 randomly taken solutions, some of which are from warmup.

## sessionInfo
```{r end, fig.width=7.2, fig.height=3.5}
print(sessionInfo())
```
